---
title: "Deaths"
author: "James Troy"
date: "`r format(Sys.time(), '%a %d %B, %Y')`"
output:
  word_document:
    toc: yes
    reference_docx: style_ref.docx
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

covid_data_url <-
  'https://opendata-geohive.hub.arcgis.com/datasets/d8eb52d56273413b84b0187a4e9117be_0.csv'
# Specify destination where file should be saved
covid_file <-
  "../data/CovidStatisticsProfileHPSCIrelandOpenData.csv"
download.file(url = covid_data_url, destfile = covid_file, 'curl')

#
# defaultW <- getOption("warn")
#
# options(warn = -1)

# options(warn = defaultW)


library("ggplot2")
library("dplyr")
library("lubridate")
library("wordcloud2")
library("stringr")
library("tidyverse")
library("webshot")
library("htmlwidgets")
library("viridis")

webshot::install_phantomjs(force = TRUE)

```

# Introduction


There has been a lot of information distributed across many platforms over the last 18 month about how deadly Covid-19 is.
The information comes from many sources, both reliable and unreliable. From Facebook to traditionally reliable media like the BBC. 

Right now people are unsure which information they believe, as governments have been caught out lying and vested interests are pushing their own agendas.

Here we will focus on Ireland and death rates in the country during the Covid-19 pandemic.

I have harvested data from [rip.ie](https://rip.ie) and propose to compare increment in deaths due to Covid-19 with normal deaths published on [rip.ie](https://rip.ie).


# The quiestion

Is the Covid-19 death rate effecting the normal national death rate?

# The Rationale

The rationale for doing this is to verify that when there is a spike in deaths from Covid-19 then there is also a spike in the national death rate.

Though we don't have access to the  actual deaths that occurred, the published deaths on rip.ie give a reasonable indication of the death rate in the Irish state.

The covid data is published by the CSO in Ireland

# Covid Deaths

##Load the data
```{r}
rip_df <- read.csv(file = '../data/deaths.csv')
```

 How many rows and columns have we in our rip data set?
 
```{r}
dim(rip_df)

```

## Prepare the date from [rip.ie](https://rip.ie)

### Remove any duplicates

For various reasons many deaths are published multiple times, but they keep the same ID. 

For example a person may be from Cork, but living in Dublin for many years. So te death may be published in Cork and Dublin.

Also a person may have a death published by more than one family, but the actual death still has a unique value.

When the undertaker publishes the death, the option to publish multiple times is offered.

```{r echo=FALSE}
rip_df <- rip_df %>%
  arrange(id, -id) %>%
  filter(duplicated(id) == FALSE)
```

#### Have we removed duplicate ID's?

```{r}
dim(rip_df)

```


#### Showing the structure of the rip dataframe
```{r echo=FALSE}
str(rip_df)
```

### Drop unapplicable columns

Remove any unnecessary columns

```{r}
drops <- c(
  "link"  ,
  "death_date",
  "maiden_name",
  "unknown_1",
  "address",
  "image",
  "id",
  "unknown_4",
  "unknown_5" ,
  "unknown_6"  ,
  "unknown_7" ,
  "unknown_8"
)

rip_df <- rip_df[,!(names(rip_df) %in% drops)]

str(rip_df)
```


### Update datetime cols


We are using the published date rather than the death date because the death date seems to after the published date.

Convert the published var as a __Date__ type

```{r}
rip_df$published <- dmy(rip_df$published)
rip_df
```


### Add a month and year column

```{r}

rip_df$Month_Yr <- format(as.Date(rip_df$published), "%Y-%m")
rip_df$month <- as.factor(month(rip_df$published , label = TRUE))
rip_df$year <- year(rip_df$published)
rip_df$week_num <- isoweek(rip_df$published)
rip_df$yday <- yday(rip_df$published)

```


### Add factors to catagorical vars
Convert some columns to factors

```{r}
rip_df$first_name <- as.factor(rip_df$first_name )
rip_df$surname <- as.factor(rip_df$surname )
rip_df$county <- as.factor(rip_df$county )
rip_df$town <- as.factor(rip_df$town )

```

### Showing the structure of the rip dataframe after some updates
```{r echo=FALSE}
str(rip_df)
```


### Summary of date from rip.ie
```{r}
rip_df[rip_df==""]<-NA 
rip_df <- rip_df %>% drop_na()
summary(rip_df)
```
## Names

Lets look at the most popular names taht passed away

Take the firstname and trim and shortened versions and make  the first character uppercase  and subsequent characters lowercase.

For example __JAMES (Jim)__ will become __James__

```{r}
trimmer <- function(x) {
  return (str_to_title(trimws(str_split(
    trimws(str_split(x, " \\(")[[1]][1]), " "
  )[[1]][1])))
}

rip_df$first_name <- lapply(rip_df$first_name, trimmer)
```


### Forenames

The most popular forenames  are traditional Irish names as expected
```{r}
first_name_count <- rip_df  %>% count(first_name)

# Make the graph
first_name_graph <-
  wordcloud2(data = first_name_count,
             color = "random-light",
             backgroundColor = "grey")
# Save png for later use
saveWidget(first_name_graph, "tmp.html", selfcontained = F)
webshot(
  "tmp.html",
  "images/first_namess.png",
  delay = 5,
  vwidth = 480,
  vheight = 480
)

first_name_graph
```


### Surnames

The most popular surnames are traditional Irish names as expected.

```{r}
set.seed(runif(1))
surname_count <- rip_df  %>% count(surname)

# Make the graph
surname_graph <-
  wordcloud2(data = surname_count,
             color = "random-light",
             backgroundColor = "grey")
# Save png for later use
saveWidget(surname_graph, "tmp.html", selfcontained = F)
webshot(
  "tmp.html",
  "images/surnames.png",
  delay = 5,
  vwidth = 480,
  vheight = 480
)
# show the graph
surname_graph

```

## Aggregating dates

Show the number of deaths published annually from 2017 to August 2021

We can see that 

- 2018 had a slight increase on 2017
- 2018 and 2019 are quite similar
- 2020 has an increase on 2019
- We are 65 % of the way through 2021 and the total so far is about 70 % of 2020

This could also be explained by increased usages in the service over time.

```{r}
by_year = rip_df %>% drop_na() %>% 
  group_by(year)  %>%
  summarise(n_deaths = n(), .groups = 'drop')

by_year %>%
  ggplot(aes(x = year, y = n_deaths)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(
    aes(label = n_deaths),
    vjust = 1.6,
    color = "white",
    size = 3.5
  ) +
  
  labs(
    title = "Published Deaths from rip.ie",
    subtitle =  paste(c(
      min(by_year$year), "to", max(by_year$year)
    ) ,
    collapse = " ") ,
    y = "Published deaths",
    x = "Year"
  )

```

We can see that January has more deaths than any other month, and supprisingly has a much bigger number than December.

My theory for this is that deaths between Christmas and the new year may have a delay in being published, hence the increase in |January
```{r}

ggplot(rip_df, aes(month, fill = year)) +
  geom_histogram(stat="count", binwidth = 500)

```




```{r}

monthly_change <- drop_na(
  rip_df %>%
    drop_na() %>%
    group_by(Month_Yr)  %>%
    summarise(n_deaths = n(), .groups = 'drop') %>%
    mutate(pct_change = ((
      n_deaths / lag(n_deaths) - 1
    ) * 100))
)
# monthly_change


ggplot(monthly_change, aes(x = Month_Yr, y = pct_change, group = 1)) +
  geom_line(color = "red", ) +
  # geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE) +
  scale_color_viridis() +

  labs(
    title = "Rate of change in deaths published from previous month",
    subtitle =  paste(c(
      min(monthly_change$Month_Yr),
      "to",
      max(monthly_change$Month_Yr)
    ) ,
    collapse = " ") ,
    y = "% change",
    x = "Year Month"
  ) +
  theme(axis.text.x = element_text(angle = 90))

```


We acn see that there are an increase in deaths published at the start of the year and a decrease 
```{r}

rip_df %>%
  
  group_by(month,year)  %>%
  summarise(n_deaths = n(), .groups = 'drop')  %>%
  ggplot( aes(x=month, y=n_deaths, group=year, color=year)) +
    geom_line() +
  scale_color_viridis() +
    ggtitle("Monthly deaths per year") 

```
##### Clean the covid data

Load the covid data and remove any unnecessry columns.
```{r}
covid_df <-
  read.csv(file = '../data/CovidStatisticsProfileHPSCIrelandOpenData.csv')
drops <- c(
  "X",
  "Y"  ,
  "HospitalisedCovidCases" ,
  "RequiringICUCovidCases" ,
  "HealthcareWorkersCovidCases",
  "ClustersNotified",
  "HospitalisedAged5"      ,
  "HospitalisedAged5to14" ,
  "HospitalisedAged15to24" ,
  "HospitalisedAged25to34" ,
  "HospitalisedAged35to44"  ,
  "HospitalisedAged45to54" ,
  "HospitalisedAged55to64"  ,
  "Male"  ,
  "Female"     ,
  "Unknown"        ,
  "Aged1to4"   ,
  "Aged5to14"   ,
  "Aged15to24" ,
  "Aged25to34" ,
  "Aged35to44"   ,
  "Aged45to54"    ,
  "Aged55to64"   ,
  "Median_Age"  ,
  "CommunityTransmission",
  "CloseContact",
  "TravelAbroad" ,
  "FID"  ,
  "StatisticsProfileDate",
  "HospitalisedAged65to74"    ,
  "HospitalisedAged75to84" ,
  "HospitalisedAged85up"  ,
  "Aged65to74"    ,
  "Aged75to84"     ,
  "Aged85up"     ,
  "week_num"    ,
  "yday"
)

covid_df <- covid_df[,!(names(covid_df) %in% drops)]

names(covid_df)
```

#### Format the dates 
```{r}
covid_df$Date <- ymd_hms(covid_df$Date)
covid_df$week_num <- isoweek(covid_df$Date)
covid_df$yday <- yday(covid_df$Date)

covid_df$Month_Yr <- format(as.Date(covid_df$Date), "%Y-%m")
covid_df$month <- as.factor(month(covid_df$Date , label = TRUE))
covid_df$year <- year(covid_df$Date)
```

#### Remove any null values
```{r}
covid_df$ConfirmedCovidDeaths[is.na(covid_df$ConfirmedCovidDeaths)] <- 0
covid_df$CovidCasesConfirmed[is.na(covid_df$CovidCasesConfirmed)] <- 0

```


#### Daily covid deaths

As we can see we have had 2 serious waves. At the start of Q2 in 2020 and Q1 in
2021 and to a lesser extent the end of Q2 in 2021

```{r}
covid_df %>%
  
  ggplot(aes(
    x = Date,
    y = ConfirmedCovidDeaths,
   
  )) +

  geom_line(color = "red", ) +
  scale_color_viridis() +
  labs(
    title = "Daily covid deaths",
    subtitle =  paste(c(
      min(covid_df$Month_Yr),
      "to",
      max(covid_df$Month_Yr)
    ) ,
    collapse = " ") ,
  ) +
  theme(axis.text.x = element_text(angle = 90))
```

#### Total covid deaths

Incremental covid deaths, again we can see 2 spikes in deaths in Q2 of 2020 and Q1 2021

```{r}

covid_df %>% ggplot(., aes(x = Date, y = TotalCovidDeaths)) +
  geom_line(color = "steelblue") +
  
  
  labs(title = "Totalcovid deaths",
       subtitle =  paste(c(
         "From",
         min(covid_df$Month_Yr),
         "to",
         max(covid_df$Month_Yr)
       ) ,
       collapse = " ") ,) +
  
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

```

#### Create deaths by date for RIP

Get the daily death count from rip.ie


```{r}
daily_rip_deaths <- drop_na( rip_df %>%
  group_by(published)  %>%
  summarise(n_rip_deaths = n(), .groups = 'drop'))
```
#### Join the dataframes

Using left join, join the covid and rip daily data

```{r}
covid_rip_join <- left_join(covid_df, daily_rip_deaths, by = c("Date" = "published"))

str(covid_rip_join)
```


#### Sum daily deaths
Add new column with accumulated deaths so far per date for RIP.ie, 
similar to TotalCovidDeaths variable

```{r}
totals = c()

for (i in 1:nrow(covid_rip_join))
  totals <- c(totals, sum(covid_rip_join$n_rip_deaths[1:i]))
covid_rip_join$totalDeaths <- totals

covid_rip_join <- covid_rip_join %>% drop_na()
```


#### Regression 

Draw a regression line between ConfirmedCovidDeaths for a day and number of 
reported rip deaths.

As the number of deaths grow it becomes less accurate.

```{r}
covid_rip_join %>%
  ggplot(aes(x = n_rip_deaths, y = ConfirmedCovidDeaths)) +
  geom_point(size=2) +
  geom_smooth(method = lm, formula = y ~ splines::bs(x, 3), se = FALSE)
```
Draw a regression line between ConfirmedCovidDeaths for a day and number of 
reported rip deaths per month.

As the number of deaths grow it becomes less accurate.

```{r}
covid_rip_join %>%
  ggplot(aes(x = n_rip_deaths, y = ConfirmedCovidDeaths)) +
  geom_point(size=2) +
  geom_smooth(method = lm, formula = y ~ splines::bs(x, 3), se = FALSE) + facet_wrap("month")

```


```{r}

covid_rip_join %>%
  ggplot(aes(x = n_rip_deaths, y = ConfirmedCovidDeaths, colour = month)) +
  geom_point(size=2) +
  geom_smooth(method = lm, formula = y ~ splines::bs(x, 3), se = FALSE)

```

```{r}
# covid_rip_join %>%
#   ggplot(aes(x = Month_Yr, y = ConfirmedCovidDeaths)) +
#   geom_point(size=2) 




covid_rip_join %>%
  gather(key, value, TotalCovidDeaths, totalDeaths) %>%
  ggplot(aes(x = Date, y = value, colour = key)) +
  geom_line() +
  labs(title = "Accumulated rip.ie and covid deaths",
       subtitle =  paste(c(
         "From",
         min(covid_df$Month_Yr),
         "to",
         max(covid_df$Month_Yr)
       ) ,
       collapse = " ") , ) +
  
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```




#### Is there a relationship?

We can see on the graph below that as the number of deaths from covid rise then
there is a spike also in published deaths from rip.ie.

```{r}


covid_rip_join %>%
  gather(key, death_count, ConfirmedCovidDeaths, n_rip_deaths) %>%
  ggplot(aes(x = Date, y = death_count, colour = key)) +
  geom_line() +
  labs(title = "Accumulated rip.ie and covid deaths",
       subtitle =  paste(c(
         "From",
         min(covid_df$Month_Yr),
         "to",
         max(covid_df$Month_Yr)
       ) ,
       collapse = " ") , ) +
  
  theme(axis.text.x = element_text(angle = 60, hjust = 1))


```










